---
title: "LoomExperiment round-trip error after logNormCounts"
author: "Kevin Rue-Albrecht"
execute:
  freeze: false  # never re-render during project render
format:
  html:
    embed-resources: true
    toc: true
    toc-location: right
---

*Generated: `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`*

## Context

This notebook illustrates a round-tripping bug related to `r BiocStyle::Biocpkg('LoomExperiment')` and `r BiocStyle::Biocpkg('scuttle')` packages.

It aims to facilitate a discussion on responsibility to determine where the fix should be applied.

## Libraries

```{r}
#| message: false
#| warning: false
library(LoomExperiment)
library(Matrix)
library(scuttle)
```

## Simulate data

Simulate a sparse count matrix (100 features x 10 barcodes).

Note: The error does not occur for a regular matrix.

```{r}
set.seed(1)
simulated_counts <- rsparsematrix(
    nrow = 100, ncol = 10, density = 1,
    rand.x = function(n) rpois(n, lambda = 5)
)
```

Create a `SingleCellLoomExperiment` from that count matrix.

Note: Reproducing the bug does not require populating other slots (e.g., `rowData`, `colData`).

```{r}
sce <- SingleCellExperiment(
    assays = list(counts = simulated_counts)
)
scle <- as(sce, "SingleCellLoomExperiment")
rownames(scle) <- paste0("feature", seq(nrow(scle)))
scle
```

## Saving and reloading the original object works

```{r}
if (!dir.exists("outputs/")) {
    dir.create("outputs")
}
if (!file.exists("outputs/scle.loom")) {
    export(scle, "outputs/scle.loom", format = "loom")
}
```

```{r}
import("outputs/scle.loom", format = "loom")
```

## Apply scuttle::logNormCounts

```{r}
scle_lognorm <- scuttle::logNormCounts(scle)
scle_lognorm
```

## The new object can be saved but not reloaded

```{r}
if (!file.exists("outputs/scle_lognorm.loom")) {
    export(scle_lognorm, "outputs/scle_lognorm.loom", format = "loom")
}
```

```{r}
tryCatch(
    import("outputs/scle_lognorm.loom", format = "loom"),
    error = function(e) conditionMessage(e)
)
```

## Temporary workaround

The code chunk below illustrates the "set dimnames of individual assays to `NULL`" workaround that got me around the issue in the [Clustering 3K PBMCs with Bioconductor](../tutorial-singlecell-bioconductor/) notebook.

```{r}
library(LoomExperiment)
library(scuttle)
sce <- import('outputs/scle.loom', format = "loom", type = "SingleCellLoomExperiment")
sce <- scuttle::logNormCounts(x = sce)
dimnames(sce@assays@data$counts) <- list(NULL, NULL)
dimnames(sce@assays@data$logcounts) <- list(NULL, NULL)
if (!file.exists("outputs/scle_lognorm_patched.loom")) {
	export(sce, "outputs/scle_lognorm_patched.loom", format = "loom")
}
import("outputs/scle_lognorm_patched.loom", format = "loom")
```

# Discussion

Quoting from Claude Code:

> `import()` doesn't eagerly load the assay matrix — it wraps the on-disk HDF5 dataset in a lazy `DelayedArray`.
> Loom files store matrices transposed (cells × genes on disk vs. genes × cells in R), so somewhere inside `import()` it calls `t(...)` on that delayed array to flip it back.
>
> To dispatch the S4 method for `t()`, R has to evaluate the argument first — which forces an actual read from the HDF5 file.
> That read fails with `HDF5. Dataset. Can't open object`, HDF5's error for "you asked me to open a dataset/group that doesn't exist." So the crash isn't in your code logic — it's `import()` trying to open something inside the file that isn't there.
>
> Every earlier loom file has symmetric (both NULL) dimnames on the assay matrix, so on import there's nothing to open.
>
> However, after scuttle::logNormCounts() runs, the resulting logcounts/counts assay matrices end up with dimnames set on only one side (length 32738 = genes), while the other side remains NULL.
>
> When export() writes that asymmetric dimnames pair, it writes only the "2" entry to disk.
>
> During re-import(), the reader apparently doesn't handle a partial dimnames group gracefully—it still tries to open both "1" and "2", and fails when "1" doesn't exist.
