---
title: "Clustering 3K PBMCs with(in) Bioconductor"
author: "Kevin Rue-Albrecht"
execute:
  freeze: true  # true: never re-render during project render
format:
  html:
    embed-resources: true
    toc: true
    toc-location: right
---

*Generated: `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`*

## Context

This tutorial demonstrates a single-cell analysis workflow using Bioconductor packages, entirely relying on R object without considering the export of data in a way that is interoperable and reusable in other software ecosystems.

## Libraries

```{r}
#| message: false
#| warning: false
library(cowplot)
library(DropletUtils)
library(scrapper)
library(scater)
library(scuttle)
```

## Data

### Data download

To create a parallel tutorial to the existing tutorial [Clustering 3K PBMCs with Seurat](https://training.galaxyproject.org/training-material/topics/single-cell/tutorials/scrna-seurat-pbmc3k/tutorial.html), input data files were downloaded from these Zenodo links:

```
https://zenodo.org/record/3581213/files/genes.tsv
https://zenodo.org/record/3581213/files/barcodes.tsv
https://zenodo.org/record/3581213/files/matrix.mtx
```

```{r}
if (!dir.exists("data")) {
	dir.create("data")
}
if (!file.exists("data/genes.tsv")) {
	download.file("https://zenodo.org/record/3581213/files/genes.tsv", "data/genes.tsv")
}
if (!file.exists("data/barcodes.tsv")) {
	download.file("https://zenodo.org/record/3581213/files/barcodes.tsv", "data/barcodes.tsv")
}
if (!file.exists("data/matrix.mtx")) {
	download.file("https://zenodo.org/record/3581213/files/matrix.mtx", "data/matrix.mtx")
}
```

### Create a SingleCellExperiment Object

The common data structure adopted by Bioconductor packages to store the matrix as well as gene and cell annotations is called `SingleCellExperiment`.

We can import the matrix and annotations of genes and cells into an `SingleCellExperiment` object using the function `read10xCounts()` of the package [DropletUtils](https://bioconductor.org/packages/DropletUtils/).

We set the `samples=` argument to the directory that contains the three 10x files.
Changing `row.names=` from the default `"id"` to `"symbol"` ensures that human-readable gene symbols are used instead of the database-friendlier Ensembl gene identifiers.
Either way, both Ensembl gene identifiers and gene symbols will be stored in the `rowData()` component of the `SingleCellExperiment` and available at all times.

```{r}
library(DropletUtils)
sce <- DropletUtils::read10xCounts(
	samples = "data/",
	row.names = "symbol"
)
```
